# Data

## Description

In this project, we will use the "Death Rates for Suicide" dataset from the CDC (Centers for Disease Control and Prevention) website. The dataset is published by the National Center for Health Statistics and is updated annually. How data is colleted is not explicitly mentioned. It is provided as a CSV file containing 21 columns and 3,786 rows. However, there are some challenges with the dataset. Likely due to privacy concerns, the dataset does not explicitly explain how the data was collected, and there is a lack of transparency regarding the calculation process for the estimates. Additionally, the dataset requires careful preprocessing for analysis, as there are hierarchical relationships between levels in some columns. For example, in the "Subgroup" column, levels such as "Female: 25-34 years" are subgroup of "Female," depending on their corresponding "Group" column level ("Sex and age group" intersects with "Sex"). This hierarchical structure necessitates detailed hard works on preprocessing. For importing the dataset, we will store it in a GitHub repository to ensure centralized and version-controlled access. Using the raw file link from GitHub, we will import the data directly into RStudio using the `read.csv()` function for analysis. Here is the link to our dataset's website containing column descriptions : https://data.cdc.gov/NCHS/DQS-Death-rates-for-suicide-by-sex-race-Hispanic-o/p7se-k3ix/about_data


## Missing value analysis

```{r, message=FALSE, warning=FALSE}
# Load necessary libraries
library(tidyverse, quietly = TRUE)
library(ggplot2)
```

```{r}
# Read the dataset
data <- read.csv("DQS_Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States_from_CDC_WONDER_20241119.csv")

# 1. Check missing values by column
missing_by_column <- colSums(is.na(data))
missing_by_column <- data.frame(Column = names(missing_by_column), Missing_Count = missing_by_column)

# 2. Check missing values by row
missing_by_row <- rowSums(is.na(data))
missing_by_row <- data.frame(Row = 1:nrow(data), Missing_Count = missing_by_row)

# 3. Visualize missingness using a simple heatmap (ggplot2)
missing_data <- data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  gather(key = "Variable", value = "Missing_Count")

ggplot(missing_data, aes(x = Variable, y = Missing_Count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip the axes for better readability
  labs(title = "Missing Data Count by Variable", x = "Variables", y = "Missing Count") +
  theme_minimal()
```

The bar chart above shows the missing value counts by variable in the dataset. We can see that there are 3 columns are all NAs towards the whoe dataset: "SUBTOPIC_ID", "SUBTOPIC" and "FLAG". One possible explanation is that this dataset doesn't have any subtopic. For columns of "STANDARD_ERROR", "ESTIMATE_UCI", "ESTIMATE_LCI" and "ESTIMATE", they have 182 missing values respectively. Judging by the name of columns, it's easy to see their correlation so they have the same counts in missing values. Notice that, "ESTIMATE" column represents the number of death per 100,000 resident population, which is the most important information in this dataset. Thus, we need dig deep in this variable. 

```{r}
missing_counts <- data %>%
  group_by(GROUP, TIME_PERIOD) %>%
  summarise(missing_count = sum(is.na(ESTIMATE)), .groups = "drop")

# Create a heatmap
ggplot(missing_counts, aes(x = GROUP, y = TIME_PERIOD, fill = missing_count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "red", na.value = "grey80") +
  labs(title = "Heatmap of Missing Values in 'ESTIMATE' by Group and Time Period",
       x = "Group",
       y = "Time Period",
       fill = "Missing Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

The heatmap reveals the distribution of missing values in the "ESTIMATE" column, across various Group categories and Time Periods. The blank spaces in the heatmap represent combinations of "GROUP" and "TIME_PERIOD" where no data exists in the dataset. This means that for certain subgroups during specific time periods, no rows are present in the data.

For the majority of groups, such as Age group, Sex, and simpler combinations like Sex and race, the data appears consistently available from 2000 to 2015. This is reflected in the lighter blue areas of the heatmap, signifying minimal missing values in these categories. These groups represent higher levels of aggregation, where data collection and reporting are typically more consistent and reliable.

However, a significant increase in missing values is observed for the year 2020 across several categories, particularly for complex subgroups such as Sex, race, and Hispanic origin, and age group. These are represented by darker shades of red or entirely blank areas in the heatmap, highlighting substantial gaps in the data. This trend suggests potential challenges in data collection or reporting for detailed subgroup intersections during this time period, which may have been influenced by disruptions such as the COVID-19 pandemic.










